--- 
title: "TF activity inference from bulk transcriptomic data with DoRothEA as 
    regulon resource."
author:
- name: Christian H. Holland 
  affiliation: Institute for Computational Biomedicine, Heidelberg University
  email: christian.holland@bioquant.uni-heidelberg.de
- name: Alberto Valdeolivas
  affiliation: Institute for Computational Biomedicine, Heidelberg University
- name: Julio Saez-Rodriguez 
  affiliation: Institute for Computational Biomedicine, Heidelberg University 
   
# package: dorothea
output:
  BiocStyle::html_document
bibliography: references.bib
abstract: |
  This vignette describes how to infer transcription factor activity from 
  bulk transcriptome data by using DoRothEA's curated regulons with viper.
license: GNU-GLPv3, please check http://www.gnu.org/licenses/
vignette: |
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Introduction

**DoRothEA** is a comprehensive resource containing a curated collection of 
transcription factors (TFs) and its transcriptional targets. The set of genes
regulated by a specific transcription factor is known as regulon. DoRothEA's 
regulons were gathered from different types of evidence. Each TF-target 
interaction is defined by a confidence level based on the number of 
supporting evidence. The confidence levels ranges from A (highest confidence) 
to E (lowest confidence) [@GarciaAlonso2019]. 
While DoRothEA was originally developed for the application on human data it can
be applied also on mouse data with comparable performace but better coverage 
than dedicated mouse regulons [@Holland2019].

**DoRothEA** regulons are usually coupled with the statistical method **VIPER** 
[@Alvarez2016]. In this context, TF activities are computed based on the mRNA 
expression levels of its targets. We therefore can consider TF activity as a 
proxy of a given transcriptional state [@Dugourd2019].
However, it is up to the user to decide which statistcal method to use. 
Alternatives could be for instance classical Gene Set Enrichment Analysis or 
simply mean statistic.

# Installation

First of all, you need a current version of R (http://www.r-project.org). In 
addition you need `r Biocpkg("dorothea")`, a freely available package deposited 
on http://bioconductor.org/ and https://github.com/saezlab/dorothea_R. 

You can install it by running the following commands on an R console:

```{r "installation", eval=FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("dorothea")
```

We also load here the packages required to run this vignette

```{r "load packages", message = FALSE}
## We load the required packages
library(dorothea)
library(bcellViper)
library(dplyr)
library(viper)
```

# Example of usage

According to the vignette from the `r Biocpkg("viper")` package we demonstrate 
how to combine viper with regulons from DoRothEA.

## Accessing example data and  DoRothEA regulons.
Similiar to the `r Biocpkg("viper")` vignette we use the gene expression matrix 
from the `r Biocpkg("bcellViper")` package. Click [here](https://bioconductor.org/packages/release/bioc/vignettes/viper/inst/doc/viper.pdf) 
for more information about the gene expression matrix. The regulons from 
DoRothEA are provided within the dorothea package and can be acessed via the 
`data()` function. As the gene expression matrix contains human data we also 
load the human version of DoRothEA.
```{r "load data"}
# accessing expression data from bcellViper
data(bcellViper, package = "bcellViper")

# acessing (human) dorothea regulons
# for mouse regulons: data(dorothea_mm, package = "dorothea")
data(dorothea_hs, package = "dorothea")
```

## Running VIPER with DoRothEA regulons

We implemented a wrapper for the viper function that can deal with different 
input types  such as matrix, dataframe, ExpressionSet or Seurat objects 
(see dedicated vignette for single-cell analysis). We subset DoRothEA to the 
confidence levels A and B to include only the high quality regulons.

```{r "run viper", message=FALSE}
regulons = dorothea_hs %>%
  filter(confidence %in% c("A", "B"))

tf_activities <- run_viper(dset, regulons, 
                           options =  list(method = "scale", minsize = 4, 
                                           eset.filter = FALSE, cores = 1, 
                                           verbose = FALSE))
```

# Session info

```{r sessionInfo, echo=FALSE}
sessionInfo()
```

# References

